### 2020-10-30 submission/manuscript.tex

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups {[}{]}. When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S6B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S6B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

@@

Removing rare sequences decreases the diversity represented by 16S rRNA
gene sequence data and increases the variation between samples. Such
impacts will hinder the statistical power to differentiate between
treatment groups. Instead of removing rare sequences, researchers should
focus on optimizing their sequence generation to minimize the amount of
PCR and sequencing errors. In addition, samples should be rarefied to a
common number of sequences across samples without prior culling of rare
sequences. The number of artifacts is correlated to the number of
sequences being considered. With this in mind, rarefaction allows one to
control for uneven sampling effort and to control for the number of
artifacts in the analysis.

Need to treat rare sequences with a grain of salt

@@

>>>>

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups (3). When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S6B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S6B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

@@

\subsubsection{Big points}\label{big-points}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Rare sequences are likely to be covered in other samples, suggesting
  they are rare; removing them biases against samples with fewer
  sequences
\item
  Removing rare sequences reduces the diversity within samples and
  changes the shape of the communities
\item
  The results include
\item
  increases the differences between samples; this makes it harder to
  detect differences between samples when those differences actually
  exist
\item
  If there is confounding between the number of reads per sample and the
  experimental grouping, then removing rare sequences increases the
  probability of falsely detecting a difference
\end{itemize}

Removing rare sequences decreases the diversity represented by 16S rRNA
gene sequence data and increases the variation between samples. Such
impacts will hinder the statistical power to differentiate between
treatment groups.

\subsubsection{What to do}\label{what-to-do}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Removal of rare sequences is largely a response to optimizing the
  number of reads and length of sequences over their quality
\item
  Researchers should instead focus on optimizing their sequence
  generation to minimize the amount of PCR and sequencing errors. We
  know that incomplete overlap of sequence reads significantly increases
  the sequencing error rate. Didn't test on crappy data
\item
  Rarefy data to control for uneven sampling of communities and PCR or
  sequencing artifacts
\end{itemize}

In addition, samples should be rarefied to a common number of sequences
across samples without prior culling of rare sequences. The number of
artifacts is correlated to the number of sequences being considered.
With this in mind, rarefaction allows one to control for uneven sampling
effort and to control for the number of artifacts in the analysis.

\subsubsection{Should we implicitly trust rare
sequences?}\label{should-we-implicitly-trust-rare-sequences}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  No.
\item
  Need to treat rare sequences with a grain of salt, but don't discard
\item
  Need to use other methods to validate results observed using rare taxa
\item
  For community metrics
\item
  rare sequencs are an important component and removing them shifts the
  community distribution.
\item
  metrics need to be treated on a relative rather than absolute basis
\end{itemize}

\subsubsection{Definition of rarity}\label{definition-of-rarity}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Idea of a blanket removal of reads below a threshold is stupid anyway
\item
  Microbial communities contain numerically rare, temporarily rare, and
  spatially rare taxa.
\item
  Removing those populations fundamentally changes how we think of
  microbial communities
\end{itemize}

@@

3. \textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss PD},
\textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
healthy nonsmokers and smokers. American Journal of Respiratory and
Critical Care Medicine \textbf{187}:1067--1075.
doi:\href{http://dx.doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.
